\bf Mainstreaming Race Science

### 2023-06-12 paper/supplement_paper.tex
readme for paper
working on main text

\author{
Daniel J. Hicks,\textsuperscript{*}
and Emilio Lobato
\\\textsuperscript{*}To whom correspondence should be addressed; E-mail:
\href{mailto:dhicks4@ucmerced.edu}{\nolinkurl{dhicks4@ucmerced.edu}}.
}

@@

\begin{document}
\maketitle
\section*{Supplementary Text}
\renewcommand{\thesection}{S\arabic{section}}
\hypertarget{materials-and-methods}{%
\section{Materials and methods}\label{materials-and-methods}}

Appendix figures and tables will also need to be moved. A basic example
is provided here so this feature is also tested.

First we have a figure.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/thrice-1.pdf}
\caption{test thrice}
\end{figure}

And next we have a table

\begin{table}[ht]
\begin{center}
\begin{tabular}{ll}
thrice & 1 \\
thrice & 2\\
\end{tabular}
\end{center}
\end{table}

>>>>

\author{
Daniel J. Hicks,\textsuperscript{*}
and Emilio Lobato
\\\textsuperscript{*}Corresponding author: Daniel J. Hicks,
\href{mailto:dhicks4@ucmerced.edu}{\nolinkurl{dhicks4@ucmerced.edu}}.
}

@@

\begin{document}
\maketitle
\section*{Supplementary Text}
\renewcommand{\thesection}{S\arabic{section}}
\hypertarget{materials-and-methods}{%
\section*{Materials and methods}\label{materials-and-methods}}

\emph{{[}probably some introduction{]}}

\hypertarget{corpus-assembly}{%
\subsection*{Corpus assembly}\label{corpus-assembly}}

Corpus assembly is summarized in figure \emph{{[}xref{]}}. The corpus
was assembled in two parts, one for \emph{Mankind Quarterly} and the
other for mainstream journals.

\emph{{[}figure{]}}

\hypertarget{mankind-quarterly}{%
\subsubsection*{Mankind Quarterly}\label{mankind-quarterly}}

An electronic archive of \emph{Mankind Quarterly}, from the first issue
in \emph{{[}1969{]}} through 2004, is available for free on the open web
from \emph{The Unz Review}, which has been characterized as a white
nationalist website by the Southern Poverty Law Center
\emph{{[}\url{https://www.splcenter.org/hatewatch/2021/01/19/meet-white-nationalist-organizer-who-spewed-hate-against-lawmakers}{]}}.
Web scraping packages and techniques \emph{{[}which?{]}} were used to
retrieve every PDF available in this archive. These PDFs were processed
using the open-source OCR (optical character recognition) software
\texttt{tesseract} and the R API \texttt{rtesseract} \emph{{[}cites{]}}.
This stage of corpus assembly was conducted in Fall 2021.

\hypertarget{mainstream-journals}{%
\subsubsection*{Mainstream journals}\label{mainstream-journals}}

To identify suitable mainstream journals for inclusion in the corpus, we
first identified academic researchers who had been funded by the Pioneer
Fund. Pioneer is an American non-profit organization founded in 1937 to
fund eugenics research \emph{{[}cite{]}}. After the \emph{Brown v. Board
of Education} ruling in 1954, Pioneer funded various segregationist
efforts across the United States. \emph{{[}Shockley and Jensen{]}}. We
reviewed a critical profile of Pioneer \emph{{[}cite{]}} as well as an
archived page from the organization's own web site \emph{{[}cite{]}},
which together listed 16 researchers who had received Pioneer funds.

We then used the author search tool in Clarivate's Web of Science
platform \emph{{[}url{]}}, retrieving publication lists for 14
researchers. These searches were conducted between 2021-09-24 and
2021-10-05 by DJH. After parsing these results, we counted how many of
the 14 researchers had published in each journal. 13 journals had
published 6 or more of the 14 Pioneer-funded researchers. We excluded
\emph{Mankind Quarterly} (as already included) as well as \emph{Science}
and \emph{Nature} (as too general) from further consideration in this
side of the corpus. 3 journals published by the American Psychological
Association (APA; \emph{American Psychologist}, \emph{Contemporary
Psychology}, and \emph{Journal of Educational Psychology}) had to be
excluded due to confusion over who could give us permission to use the
archives for a text mining project, with both APA and ProQuest asserting
that we needed to get permission from the other entity. \emph{European
Journal of Personality}, published by SAGE, also had to be excluded
because our institutional access only went back to 1999.

The remaining 5 journals are all published by major academic publishers
--- Elsevier, Springer, or Cambridge University Press --- and each item
in the entire run of each journal has been assigned a DOI (digital
object identifier) for archival purposes. We used the Crossref API and
\texttt{rcrossref} R interface to this API \emph{{[}cite{]}} to retrieve
metadata for each item published from 1960-2010 in each of these 6
journals. These metadata included item-level license information ---
confirming that the text of each item could be used for text mining
projects --- and a URL to an electronic version of the item. These URLs
were used to retrieve an HTML or PDF version of each item, except for
\emph{Personality and Individual Differences}. This journal has
published a relatively large number of non-article documents, such as
book reviews and commentaries, which are not available at the URL
included in the Crossref metadata. (This is unfortunate, as it was not
difficult to find highly relevant documents that we could not
automatically retrieve and therefore excluded from our corpus
\emph{{[}review{]}}.) Instead we used Elsevier's ScienceDirect API
\emph{{[}url{]}} to search and retrieve all items from \emph{Personality
and Individual Differences} that are available. PDFs were run through
OCR and text extraction as necessary, as with the \emph{Mankind
Quarterly} documents.

All together, \emph{{[}corpus summary, total and by journal; table{]}}

\hypertarget{data-preparation}{%
\subsection*{Data preparation}\label{data-preparation}}

After document retrieval and text extraction, we pre-processed the text
using the \texttt{spaCy} NLP (natural language processing) Python
library and the R API \texttt{spacyr} \emph{{[}version, cites{]}}.
Specifically, we applied regular expressions to remove header/footer
copyright notices and hyphenation, used spaCy to annotate and extract
noun phrases (eg, ``the intelligence test items''), and then cleaned and
standardized these phrases (eg, removing the/an/a, coverting all text to
lowercase, and replacing all whitespace with underscores:
``intelligence\_test\_items''). We then counted the occurrence of each
noun phrase in the document. The aggregated ``document-term matrix'' was
stored in Parquet format \emph{{[}cite{]}} for performance reasons, and
written and read using the \texttt{arrow} package for R
\emph{{[}cite{]}}. \emph{{[}availability{]}}

NLP-extracted noun phrases offer a number of advantages over the more
traditional unigram (``single-word'') tokens. First, noun phrase
extraction removes many standard stopwords (articles, common verbs)
without relying on a fixed, a priori stopword list. Phrases can be more
informative than single terms, for example, distinguishing
``intelligence test'' from ``hypothesis test.'' But simple n-gram
extraction will include numerous phrases that are not especially
meaningful. Consider the sentence ``Since our first analyses of feeding
patterns in rats, we had been using a criterion of 40 minutes (Le Magnen
\& Tallon 1963; 1966)'' *{[}10.1017\_\_\^{}\_\_s0140525x0000042x{]}*.
Bigrams such as ``since our'' and ``criterion 40'' will likely be
discarded in vocabulary selection, but significantly increase the
computational cost of vocabulary selection. Noun phrase extraction is
therefore more efficient.

After noun phrase extraction, the corpus comprised \emph{{[}tokens{]}}
of \emph{{[}distinct noun phrases{]}}.

\hypertarget{vocabulary-selection}{%
\subsubsection*{Vocabulary selection}\label{vocabulary-selection}}

We took an information-theoretic approach to vocabulary selection.
Consider a game in which I draw a document from the corpus, then a
single token from that document. I tell you the term, and you have to
guess which document I picked. Intuitively, highly informative terms (in
this project, noun phrase types) are distinctive, allowing you to
dramatically narrow down the list of potential documents. This
``informativeness'' of a term can be quantified as the KL
(Kullback-Leibler) divergence from a ``baseline'' distribution of
documents to the distribution conditional on the term. Because the most
informative terms tend to be typos and OCR errors --- these are unique
to a single document --- we multiply the KL divergence by the logarithm
of the total number of occurrences of the term across the entire corpus.
We refer to the resulting measure as \(log(n) \Delta H\) or simply
\texttt{ndH}.

\emph{{[}Hicks 2021?{]}} used the \texttt{ndH} approach in a topic
modeling analysis, where the baseline distribution was the uniform
distribution across documents. In the current project, we found that
this approach heavily favored recurrent noun phrases in the longest
documents. Many of the documents published in \emph{Psychological
Reports} are extremely short, 1-2 reports of a single study; while many
of the documents published in \emph{Brain and Behavioral Sciences} are
book-length collections that include a long review article and sometimes
dozens of commentaries. Very generic noun phrases that happen to appear
in the latter can occur orders of magnitude more often than highly
distinctive phrases in the former, and so the \(log(n)\) factor
overwhelms the \(\Delta H\) factor.

To address this, we used a different baseline distribution of documents,
namely, one in which document probability is proportional to length.
This makes phrases from short documents much more ``surprising'' (much
less likely to occur according to the baseline), and hence substantially
increases their informativeness. This was more effective at identifying
useful phrases from across the corpus.

A common rule of thumb in topic modeling is that the vocabulary should
have about 10 times as many distinct terms as the number of documents in
the corpus \emph{{[}cite{]}}. However, we had some concerns with
computational demands here: the resulting document-term matrix would
have roughly \(10 \times 33,000^2\) or 10.9 billion entries; with 10\%
density this would require on the order of 4 GB of memory just for a
single copy of the matrix; and most of the analysis was to be conducted
on the authors' laptops. We therefore chose to work with three smaller
vocabularies, \(5 \times\), \(1 \times\) and \(\frac{1/5} \times\) the
number of documents. We refer to these as the ``large,'' ``medium,'' and
``small'' vocabulary, respectively, and they include \emph{{[}counts{]}}
distinct phrases. We compare findings across vocabularies as a
robustness check.

\hypertarget{topic-modeling}{%
\subsection*{Topic modeling}\label{topic-modeling}}

To fit topic models, we followed the approach proposed by
\emph{{[}cite{]}}, which uses varimax-rotated partial principal
components instead of the variational inference methods used by standard
topic model packages \emph{{[}eg, stm{]}}. This novel approach was
implemented in the R package \texttt{tmfast}, \emph{{[}cite{]}}. A
simulation study of \texttt{tmfast} found that it was significantly
faster and only slightly less accurate at reconstructing known
word-topic and topic-document distributions, compared to the standard
topic modeling package \texttt{stm}.

Topic models were fit for all three vocabularies (large, medium, and
small) with \(k = 5, 10, 20, 30, 40, 50\) (number of topics), resulting
in a total of 18 models.

Following the approach of \emph{{[}Hicks 2021?{]}}, we did not attempt
to identify a unique best model for further analysis. While we focus on
the medium vocabulary, \(k=30\) model as the ``median'' among the 18
models fit, we compare and contrast findings from this model with those
from the other models.

\hypertarget{topic-model-interpretation}{%
\subsubsection*{Topic model
interpretation}\label{topic-model-interpretation}}

After fitting topic models, our first research question was whether we
could identify distinctive ``race science'' topics. We addressed this
question in two ways. Following common practice in topic modeling, we
extracted lists of the top \(n\) (highest-probability) terms (noun
phrases) from each topic in each model, with \(n\) generally ranging
between 5 and 15 as we explored the models. \emph{{[}tables + Silge
plots{]}}

We also adapted the ``discursive space'' analysis proposed by
\emph{{[}Hicks 2021?{]}}. This analysis combines the Hellinger distance
between topic-document distributions --- capturing how ``far apart'' two
documents are --- with dimensional reduction techniques used for data
visualization. This produces a two-dimensional visualization of
individual documents, arranged so that documents with similar topic
distributions are visually clustered and visually distinct from
documents with different topic distributions. \emph{{[}availability{]}}

Specifically, for each fitted topic model, we extracted topic-document
distributions for each document and calculated pairwise Hellinger
distances. We then processed these distances using the UMAP algorithm
for dimension reduction, as implemented in the \texttt{umap} package
\emph{{[}details; cite{]}}. To reduce computational requirements for
these steps, we first dropped documents published in the journal
\emph{{[}Psychological Reports{]}} (\emph{{[}x documents removed, y
remaining{]}}; note that the full corpus was used to fit the topic
models, \emph{Psychological Reports} was removed just for this
visualization). Using the \texttt{plotly} library \emph{{[}cite{]}}, we
construct an interactive visualization: each point represents a single
document, and points can be colored based on either their journal or
highest-probability topic. Tooltips are displayed when we hover over a
point, including document metadata (title, journal, authors, year) as
well as ``key phrases.''

To calculate key phrases for a given document \(d\), we first identified
the highest-probability topic \(t\) for that document and the word-topic
(noun phrase-topic) distribution \(\beta_{w,t}\) for that topic. We also
retrieved the document-term matrix, containing the count \(n_{w,d}\) of
occurrences for each word (noun phrase), and calculated a weighting
\(n_{w,d} \beta_{w,t}\) for each word. The key phrases are the 5 most
highly-weighted terms for the given document.

\hypertarget{topic-quality-assessment}{%
\subsubsection*{Topic quality
assessment}\label{topic-quality-assessment}}

We next conducted a topic quality check of topic 19 from the medium
vocabulary, \(k=30\) model. Keyword and inspection of the UMAP
interactives seemed to indicate that this topic was a ``race science
discourse'' topic in this model.

To confirm this initial impression, a spreadsheet of all articles with
\(\gamma > 0.25\) for this topic was extracted (excluding articles
published in \emph{Psychological Reports}), and the top 200 articles
were reviewed manually by both authors. For this qualitative analysis,
we defined \emph{scientific racism} as (purporting) to justify racial
inequality and colonialism by appealing to the epistemic authority of
science; \emph{race science} as (pseudo-)scientific research that can be
utilized for scientific racism; and \emph{race science discourse} as
treating race science as a legitimate area of scientific research; this
includes methodological critiques of race science and empirical tests
that falsify race science hypotheses. Our qualitative review of the
top-200 topic-19 articles focused on classifying them as \emph{race
science discourse}.

For the first round of review, both authors worked independently,
dichotomously coding each article as race science discourse or not. We
calculated the ``false positive rate'' --- documents with a very high
\(\gamma\) in this topic that were not race science discourse --- and
interrater reliability using Cohen's \(\kappa\).\\
\emph{{[}false positive rate{]}} \emph{{[}Initial interrater reliability
was low, \(\kappa = 0.4\), with 50 out of 200 articles classified
discordantly.{]}} To understand the disagreement, both authors examined
the 50 discordant articles using an interactive grounded theory
approach. After this process, some initial codings were changed and
\emph{{[}almost{]}} all of the remaining \emph{{[}n{]}} discordant
articles were coded as follows:

\begin{description}
\tightlist
\item[hereditarianism and/or eugenics]
29 papers. On the ``eugenics'' side, mostly examines or otherwise
discusses the ``dysgenic'' claim that ``low-intelligence'' people have
more children than ``high-intelligence'' people. The ``hereditarianism''
side is more varied, with papers on environmentalist interventions to
raise IQ, genetics of IQ, and several other sub-topics.
\item[Flynn effect]
15 papers. Supporting, challenging, or otherwise focusing on Flynn's
findings that IQ has increased over time.
\item[Lynn/national IQ]
3 papers. Supports, challenges, or focuses on international IQ
data/comparisons made by Richard Lynn.
\end{description}

We agreed that, for all three categories of papers, there was reasonable
ambiguity about whether they should be classified as race science
discourse. These papers either do not explicitly mention race, or do so
with just a brief mention of the ``race and intelligence controversy.''
On the other hand, they either contribute to or legitimatize areas of
research that are deeply entangled with race science, and might thereby
contribute to public perceptions of a legitimate scientific debate about
race and intelligence.

Based on this qualitative review, we judged that the topic model
approach was sufficiently sensitive (low ``false positive'' rate) in
identifying race science discourse.

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-BarkanRetreatScientificRacism1992}{}}%
Barkan, Elazar. 1992. \emph{The {Retreat} of {Scientific Racism}:
{Changing Concepts} of {Race} in {Britain} and the {United States
Between} the {World Wars}}. {Cambridge University Press}.

\leavevmode\vadjust pre{\hypertarget{ref-BeagleholeRaceQuestion1950}{}}%
Beaglehole, Ernest, Juan Comas, L. A. Costa Pinto, Franklin Frazier,
Morris Ginsberg, Humayun Kabir, Claude Levi-Strauss, and Ashley Montagu.
1950. {``The {Race Question}.''} \emph{UNESCO and Its Programme} 1 (3).

\leavevmode\vadjust pre{\hypertarget{ref-BrattainRaceRacismAntiracism2007}{}}%
Brattain, M. 2007. {``Race, {Racism}, and {Antiracism}: {UNESCO} and the
{Politics} of {Presenting Science} to the {Postwar Public}.''} \emph{The
American Historical Review} 112 (5): 1386--1413.
\url{https://doi.org/10.1086/ahr.112.5.1386}.

\leavevmode\vadjust pre{\hypertarget{ref-MeadComingAgeSamoa1928}{}}%
Mead, Margaret. 1928. \emph{Coming of {Age} in {Samoa}}. {New York}:
{William Morrow and Company}.

\leavevmode\vadjust pre{\hypertarget{ref-MehlerFoundationFascismNew1989}{}}%
Mehler, Barry. 1989. {``Foundation for Fascism: {The} New Eugenics
Movement in the {United States}.''} \emph{Patterns of Prejudice} 23 (4):
17--25. \url{https://doi.org/10.1080/0031322X.1989.9970026}.

\leavevmode\vadjust pre{\hypertarget{ref-SlobodianUnequalMindHow2023}{}}%
Slobodian, Quinn. 2023. {``The {Unequal Mind}: {How Charles Murray} and
{Neoliberal Think Tanks Revived IQ}.''} \emph{Capitalism: A Journal of
History and Economics} 4 (1): 73--108.

\end{CSLReferences}
